Exploring environmental impacts of colonial contacts on Indigenous communities in northeastern Taiwan

### 2025-02-27 analysis/paper/paper.qmd
work on discussion section

Only the faunal assemblages identified to at least the family level were considered in this study for consistent comparison across time sequences. Deer antlers were excluded since they were less likely related to dietary practices. To obtain generalized pattern of resource use, measures of ubiquity and taxonomic abundance were applied. Ubiquity measures the frequency of specific taxa through their presence or absence in particular depositional contexts, suggesting the regularity of animal consumption [@Mckechnie2016; @Vanderwarker2010]. Here ubiquity values are expressed as the percentage of units with a given taxon to the total number of units per cultural layers. Faunal assemblages were primarily quantified using the Number of Identified Specimens (NISP) method, with the Minimum Number of Individuals (MNI) applied when necessary, such as for estimating the expected NISP at a site [@Grayson2004; @Lyman2012]. Different from the minimum number of individuals (MNI) that determines the actual numbers of animals recovered from remains, NISP offers a broad overview of dietary practices by considering all fragmented bone assemblages. Despite the criticism of overestimation the number of animals, NISP captures better variation in species abundance on a site with less impacts from sample size [@Grayson1984; @Lyman2008]. For comparison of identified taxa across classes, the normalized NISP were used to avoid potential biases in identification due to differences in the preservation or recovery of skeletal elements [@Grayson2004]. The normalized NISP is calculated by dividing NISP by the number of skeletal elements.

@@

fauna_combined_context_all <-
  rbind(fauna_sam, fauna_H_sam) %>% # combine the two datasets
  filter(!`部位/名稱` %in% c("角","犄角","角?")) %>%  #not associated with diet
  filter(!`部位/名稱`== "角基部"|!is.na(`部位/左右`)) %>% # associated with cranial parts
  mutate(`重量(g)` = as.numeric(`重量(g)`)) %>%
  group_by(taxa) %>%
  mutate(`Weight (g)` = sum(`重量(g)`, na.rm = T)) %>%
  ungroup() %>%
  mutate(period = factor(period, levels = c("CL1","CL2","CL3","CL4","CL5","CL6"), order = T)) %>%
  mutate(category = case_when(str_detect(animal, "sika")|str_detect(animal, "sambar")|
                              str_detect(animal, "e deer")~ "deer", str_detect(animal, "fish") ~ "fish",
                              str_detect(animal, "cattle")|str_detect(animal, "buffalo")~ "cattle",
                              animal == "aves" ~ "bird", TRUE ~ animal)) %>%
  mutate(class = case_when(str_detect(taxa, "muntjac")|str_detect(taxa, "ammal")|str_detect(taxa, "erv")|
                           str_detect(taxa, "Sus")|str_detect(taxa, "Rattus")|str_detect(taxa, "Rusa")|
                           str_detect(taxa, "Bos")|str_detect(taxa, "Bubalus")~ "mammal",
                           str_detect(category, "fish") ~ "fish", str_detect(category, "bird") ~ "bird",
                           TRUE ~ "reptile")) %>%
  mutate(`部位/左右` = case_when(`部位/左右` == "Ｌ" ~ "L", `部位/左右` == "Ｒ" ~ "R", TRUE ~ `部位/左右`)) %>%
  mutate(category = fct_relevel(category, "deer","boar","muntjac","cattle","rat","bird","fish","turtle")) %>%
  mutate(cutmarks = ifelse(str_detect(`人為痕跡`, "切"),"yes", "no"),
         fractures = ifelse(str_detect(`人為痕跡`, "削")|str_detect(`人為痕跡`, "砍")|
                            str_detect(`人為痕跡`, "折"), "yes", "no"),
         burning =  ifelse(str_detect(`人為痕跡`, "火")|str_detect(`人為痕跡`, "燒"), "yes", "no")) %>%
  filter(is.na(refitted)) # for NISP

@@

@@

Since deer is the most frequent animal at Kiwulan, the frequency of skeletal parts was examined to explore any changes in processing decisions of meat portions between different phases. Deer bones were grouped into six body portions based on their nutritional values [@Madrigal2004], including head (cranium, mandible), axial (vertebrae, sternum, ribs), upper forelimb (scapula, humerus, ulna, and radius), upper hindlimb (femur, tibia, innominate), lower limb (carpals, metacarpals, tarsals, metatarsals, astragalus, calcaneus), and foot (phalanges). The result (@fig-deer-portion: A) shows that the meaty portion, especially the upper forelimb, is the most frequent part from CL6 to CL2, followed by upper hindlimb and the less-meaty portions, lower limb and foot. The consistency of higher numbers of upper limb and lower limb suggests that those elements should be transported together (e.g., @Stiner1991). The genereally low frequency of axial element, such as vertebrae and ribs, might be likely due to a high degree of fragmentation, butchery practices, or dietary preferences. A different pattern between layers is observed in the frequency of cranial bones, which were present in CL5, increased to the highest number in CL4, and then decreased thereafter. None of cranial bones were found in CL6 and CL2. In addition, lower limb and foot bones appear to be proportionally less in CL2 and CL3 compared to earlier periods.

Considering the impacted of differential bone density on preservation of skeletal elements, the relationship between the survivorship of deer skeletal elements and bone mineral density was tested by Pearson correlation. The values of bone survivorship were collected from @Lyman1994's research on deer, and calculated by dividing NISP by expected NISP, followed by multiplying 100%. The result (@fig-deer-portion: B) shows that there is no significant positive correlation ($r$ = `r survivor_cor[1]`, P = `r survivor_cor[2]`)) between bone survivorship and bone density, suggesting that density-mediated bias has little impact on the taphonomic processes of deer bones.

@@

```{r}
#| label: tbl-deer-modify
#| echo: false
#| tbl-cap: "Number and proportion of deer bones with surface modifications at Kiwulan."
#| tbl-cap-location: bottom

@@

Faunal remains recovered from the layer associated with the European presence reveals the highest quantity, followed by the pre-European period. During the European period, in addition to the predominance of deer, an increased proportion of boar among faunal remains suggests a slightly greater reliance on boar as a secondary meat source compared to other time periods, along with a broader range of medium and small animals, such as muntjac, birds, fish, and reptiles. A similar pattern was observed in the ubiquity of animals, which shows the highest ubiquity of deer, boar and munjact remains during the European period. Interestingly, deer bones from the pre-European period, specifically CL5, presented the same ubiquity as in the European period but in only half the quantity. This might indicate consistency in the frequency and distribution of deer at the settlement scale before and after European presence, but with double the quantity of deer consumed during the European period. During the Chinese period, the low quantity of deer in CL2 and the predomination of cattle in CL1 suggest a sudden shift in dietary practices after the Chinese arrival, indicating a Chinese influence and the decline of indigeonus population in this region.

The skeletal parts of deer demonstrate a different

>>>>

Only the faunal assemblages identified to at least the family level were considered in this study for consistent comparison across time sequences. Deer antlers without attaching to crania were excluded since they were less likely related to dietary practices. To obtain generalized pattern of resource use, measures of ubiquity and taxonomic abundance were applied. Ubiquity measures the frequency of specific taxa through their presence or absence in particular depositional contexts, suggesting the regularity of animal consumption [@Mckechnie2016; @Vanderwarker2010]. Here ubiquity values are expressed as the percentage of units with a given taxon to the total number of units per cultural layers. Faunal assemblages were primarily quantified using the Number of Identified Specimens (NISP) method, with the Minimum Number of Individuals (MNI) applied when necessary, such as for estimating the expected NISP at a site [@Grayson2004; @Lyman2012]. Different from the minimum number of individuals (MNI) that determines the actual numbers of animals recovered from remains, NISP offers a broad overview of dietary practices by considering all fragmented bone assemblages. Despite the criticism of overestimation the number of animals, NISP captures better variation in species abundance on a site with less impacts from sample size [@Grayson1984; @Lyman2008]. For comparison of identified taxa across classes, the normalized NISP were used to avoid potential biases in identification due to differences in the preservation or recovery of skeletal elements [@Grayson2004]. The normalized NISP is calculated by dividing NISP by the number of skeletal elements.

@@

fauna_combined_context_all <-
  rbind(fauna_sam, fauna_H_sam) %>% # combine the two datasets
  filter(!`部位/名稱` %in% c("角","犄角","角?")) %>%  #not associated with diet
  filter(!`部位/名稱`== "角基部"|!is.na(`部位/左右`)) %>% # associated with cranial parts
  mutate(`重量(g)` = as.numeric(`重量(g)`)) %>%
  group_by(taxa) %>%
  mutate(`Weight (g)` = sum(`重量(g)`, na.rm = T)) %>%
  ungroup() %>%
  mutate(period = factor(period, levels = c("CL1","CL2","CL3","CL4","CL5","CL6"), order = T)) %>%
  mutate(category = case_when(str_detect(animal, "sika")|str_detect(animal, "sambar")|
                              str_detect(animal, "e deer")~ "deer", str_detect(animal, "fish") ~ "fish",
                              str_detect(animal, "cattle")|str_detect(animal, "buffalo")~ "cattle",
                              animal == "aves" ~ "bird", TRUE ~ animal)) %>%
  mutate(class = case_when(str_detect(taxa, "muntjac")|str_detect(taxa, "ammal")|str_detect(taxa, "erv")|
                           str_detect(taxa, "Sus")|str_detect(taxa, "Rattus")|str_detect(taxa, "Rusa")|
                           str_detect(taxa, "Bos")|str_detect(taxa, "Bubalus")~ "mammal",
                           str_detect(category, "fish") ~ "fish", str_detect(category, "bird") ~ "bird",
                           TRUE ~ "reptile")) %>%
  mutate(`部位/左右` = case_when(`部位/左右` == "Ｌ" ~ "L", `部位/左右` == "Ｒ" ~ "R", TRUE ~ `部位/左右`)) %>%
  mutate(category = fct_relevel(category, "deer","boar","muntjac","cattle","rat","bird","fish","turtle")) %>%
  mutate(cutmarks = ifelse(str_detect(`人為痕跡`, "切"), "yes", "no"),
         fractures = ifelse(str_detect(`人為痕跡`, "削")|str_detect(`人為痕跡`, "砍")|
                            str_detect(`人為痕跡`, "折")|str_detect(`人為痕跡`, "敲"), "yes", "no"),
         burning =  ifelse(str_detect(`人為痕跡`, "火")|str_detect(`人為痕跡`, "燒"), "yes", "no")) %>%
  filter(is.na(refitted)) # for NISP

@@

# calculate MNE
deer_MNE <-
  fauna_deer_only  %>%
  filter(!is.na(`部位/左右`)&is.na(overlap)) %>% # remove overlapped fragments
  count(period,`部位/左右`,`部位/名稱`,`部位/位置`) %>%
  group_by(period) %>%
  summarise(MNE = sum(n), across())

deer_NISP <- fauna_deer_only %>% count(period) %>% rename(NISP = n)
frag_index <- left_join(deer_MNE, deer_NISP) %>% mutate(index = MNE/NISP) %>% distinct(period, index) %>% 
  mutate(index = round(index, 2))
frag_CL3_CL4 <- c(frag_index$index[2], frag_index$index[3])

@@

Since deer is the most frequent animal at Kiwulan, the frequency of skeletal parts was examined to explore any changes in processing decisions of meat portions between different phases. Deer bones were grouped into six body portions based on their nutritional values [@Madrigal2004], including head (cranium, mandible), axial (vertebrae, sternum, ribs), upper forelimb (scapula, humerus, ulna, and radius), upper hindlimb (femur, tibia, innominate), lower limb (carpals, metacarpals, tarsals, metatarsals, astragalus, calcaneus), and foot (phalanges). The result (@fig-deer-portion: A) shows that the meaty portion, especially the upper forelimb, is the most frequent part from CL6 to CL2, followed by upper hindlimb and the less-meaty portions, lower limb and foot. The consistency of higher numbers of upper limb and lower limb suggests that those elements should be transported together (e.g., @Stiner1991). The genereally low frequency of axial element, such as vertebrae and ribs, might be likely due to a high degree of fragmentation, butchery practices, or dietary preferences. A different pattern between layers is observed in the frequency of cranial bones, for which a few present in CL6, increased to the highest number in CL4, and then decreased thereafter. In addition, lower limb and foot bones appear to be proportionally less in CL2 and CL3 compared to earlier periods.

Considering the impacted of differential bone density on preservation of skeletal elements, the relationship between the survivorship of deer skeletal elements and bone mineral density was tested by Pearson correlation. The values of bone survivorship were collected from @Lyman1994's research on deer, and calculated by dividing NISP by expected NISP, followed by multiplying 100%. The result (@fig-deer-portion: B) shows that there is no significant positive correlation ($r$ = `r survivor_cor[1]`, P = `r survivor_cor[2]`)) between bone survivorship and bone density, suggesting that density-mediated bias has little impact on the taphonomic processes of deer bones. For the degree of bone fragmentation, CL3 shows higher fragmentation (index = `r frag_CL3_CL4[1]`), while bones in CL4 is less fragmented (index = `r frag_CL3_CL4[2]`). However, the fragmentation rates across all layers are not significantly differed.

@@

```{r}
#| label: tbl-deer-modify
#| echo: false
#| tbl-cap: "Number and proportion of deer bones with surface modifications at Kiwulan."
#| tbl-cap-location: bottom
#| tbl-colwidths: [10,20,20,20,20,10]

@@

Faunal remains recovered from the layer associated with the European presence reveals the highest quantity, followed by the pre-European period. During the European period, in addition to the predominance of deer, an increased proportion of boar among faunal remains suggests a slightly greater reliance on boar as a secondary meat source compared to other time periods, along with a broader range of medium and small animals, such as muntjac, birds, fish, and reptiles. A similar pattern was observed in the ubiquity of animals, which shows the highest ubiquity of deer, boar and munjact remains during the European period. Interestingly, deer bones from the pre-European period, specifically CL5, presented the same ubiquity as in the European period but in only half the quantity. This might indicate consistency in the frequency and distribution of deer at the settlement scale before and after European presence, but with double the quantity of deer consumed during the European period. During the Chinese period, the low quantity of deer in CL2 and the predominate of cattle in CL1 suggest a sudden shift in dietary practices after the Chinese arrival, indicating a Chinese influence and the decline of indigenous population in this region.

The skeletal part frequencies of deer demonstrate more complete carcasses processed on site during the European period and pre-European period compared to post-European period and Chinese period. The whole carcasses processing and a substantial rise in head parts (eg. crania, mandibles) in the European period align with its high quantity of deer, suggesting more deer were brought back intact to the settlement rather than selectively butchered in the field. On the other hand, the low frequency of head parts in the earliest layer and Chinese period might indicate deer could be processed at hunting location with mostly meaty (eg. upperlimb) and mid-meaty portions (eg. ribs) transported back. Interestingly, front quarters (eg. humeri, radius) are more frequently found compared to forequarters (eg. femur, tibia) among the meaty portions. In addition to culinary preferences or differential processing, a likely explanation for this is that femur and tibia were heavily fragmented due to marrow extraction for their high yields [@Madrigal2002], which could lead to underrepresented resulting from unidentifiable small fragments.
